Power simulation section

### 2023-08-10 docs/manuscript.Rmd
revamped intro and knitted pdf

abstract: |
  Understanding how humans interact with articifical intelligence through experiments becomes increasingly relevant. However, human-AI interaction researchers lack the appropriate tools to conduct power and data analyses for the required complex study designs. In this work, we provide a tutorial on how to run customized power analyses using data simulation based on Generalized Linear Mixed Models (GLMMs). By providing code in a case study, we equip human-AI interaction researchers with the tools to simulate their own data and run analyses based on GLMMs. We discuss implications and give an outlook.

@@

<!-- General intro, relevance of Human AI Interaction experiments, problem of underpowered & inappropriately analyzed work (do not shit too much on prior research :D) -->
As the integration of artificial intelligence (AI) into our daily lives continues to grow, it has become increasingly important to study how humans interact with these intelligent systems.This is particularly important as AI becomes more sophisticated and is used in decision-making processes, for instance in medicine, that have significant impacts on individuals and society as a whole. Therefore, thorough experiments are needed to study human-AI interaction to understand how people perceive and respond to these novel technologies. When conducting experimental research in this emerging field, it is essential to determine an appropriate sample size to ensure that the results obtained are both statistically significant and reliable. Therefore, running sample size estimations is crucial for designing experiments that can provide meaningful insights into human-AI interaction.

<!-- Sample size planning is increasingly important! -->
In empirical research relying on hypothesis testing, the most common strategy to determine appropriate samples size is based on statistical power.
Statistical power is defined as the probability that a hypothesis test of interest turns out significant, when analyzing repeated samples from a population with a true effect of some pre-specified size.
Less formally, power is also described as the probability that a hypothesis test accepts the alternative hypothesis $H_1$ if $H_1$ is indeed true.
Power analysis (i.e. calculating the statistical power for a given true effect and sample size) plays a crucial role in addressing the replication crisis and meeting the requirements set forth by journals and funding agencies in human-AI interaction research.
The replication crisis has underscored the problem of high false discovery rates (FDR) coupled with low-powered studies.
Power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes needed to achieve sufficient statistical power for reliable results.
Moreover, many journals and funding agencies now mandate the inclusion of power analysis as part of study protocols and grant proposals, recognizing their significance in ensuring robust and meaningful findings.
By incorporating power analysis into research planning, researchers can enhance the replicability and credibility of their work, ultimately contributing to the advancement of psychological science.

## Customized power simulations are hard

<!-- Power simulations are hard! (especially for GLMMs) -->
Power analysis can be challenging, particularly when analyzing complex study designs using (generalized) linear mixed models (GLMMs; also called multilevel models).
While user-friendly software is available for simple statistical models like t-tests, ANOVA, and linear regression, the increasing use of GLMMs in human-AI interaction research introduces greater complexity to power analysis.
For GLMMs, analytic options to derive sample sizes are not available for general research designs.
In contrast, simulations are often required, necessitating assumptions for all parameters in the model.
An increasing number of tutorials describe how to perform power simulation for multilevel models.
However, most tutorials discuss only linear mixed models (LMMs) and focus on the most common designs, limiting the guidance available for researchers facing more complex study designs [REF].
Moreover, existing tutorials often rely on heuristics for specifying variance components (e.g. standard deviation of random intercepts) or assume that meta-analytic results or data from pilot studies is available to determine plausible values for all model parameters.

<!-- Research Gap & Summary -->
When planning a human-AI interaction experiment with complexities such as the use of GLMMs, non-standard designs, and missing by design conditions, the process of power simulations becomes more challenging.
Currently available software packages for power simulations do not include common designs in human-AI interaction research, making it necessary to build data simulations tailored specifically to the study design.
Unfortunately, there are no general recommendations that can be applied universally in such cases.
Instead, researchers must acquire the necessary skills based on the specific application at hand.
In this paper, we present a case study that serves as a practical demonstration of how to perform a data simulation for a concrete study for which no meta-analytic results or conclusive data from pilot studies is available.
Through this case study, we aim to provide guidance and insights into the complexities involved in addressing complex design considerations and obtaining accurate power estimates.

@@

3.  Teaching Bespoke Power Simulation Skills to Human-AI Interaction Researchers: The ability to conduct bespoke power simulations is a valuable skill that should be taught to Human-AI interaction researchers. By incorporating this training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power and improve the quality of their experimental designs. Equipping Human-AI interaction researchers with the knowledge and tools to perform bespoke power simulations empowers them to make informed decisions and enhance the rigor of their studies.

@@

5.  Recognizing the Value of Bespoke Power Simulations: Bespoke power simulations are not mere technicalities; they are valuable research contributions that deserve recognition in Human-AI interaction research. They offer insights into the reliability and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power. The importance of bespoke power simulations can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports.

@@

In our case study, we have performed power simulations based on a single set of parameter values that reflect our assumptions of an expected effect size.
Instead of extracting this expected effect size from meta-analyses or pilot data which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge which can be considered a vague theoretical model about the data-generating process.
When sample sizes are chosen based on the results of our power simulations, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present.
However, if the true effect is indeed smaller, power will be lower the study might not be sufficiently informative.
A common, more conservative strategy for sample size justification is to perform power analysis for a smallest effect size of interest (SESOI). 
An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero.
For strategies on the even more difficult task of specifying a plausible SESOI as well as a thorough discussion of various topics concerning power analysis, see @lakensImprovingYourStatistical2022.
When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power for different effect sizes in what is called *sensitivity power analysis*.
By simulating power for different effect sizes (in addition to the different number of subjects and items), one can make sure that power would still be sufficient to detect smaller effect sizes than our expected effect, or at least get an impression how strongly power depends on the size of the true effect.
In simple study designs, it is possible to perform sensitivity power analysis based on a single standardized effect size (e.g. analyse power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8).
However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions.
In this setting, sensitivity power analysis would require to manually specify additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question.
Power could then be simulated for several of these scenarios (across different number of subjects and items as considered earlier).

@@

Bespoke power simulations play a critical role in human-AI interaction research, allowing researchers to tailor power estimation to the unique aspects of their experiments. The skills required to perform these simulations should be taught to Human-AI interaction researchers, fostering a deeper understanding of statistical power and enhancing research design. It is essential to bridge the gap between perceived and actual effort associated with power simulations and recognize their value as research contributions. By integrating bespoke power simulations with open science and preregistration practices, researchers can improve the robustness and transparency of their findings, advancing the field.

>>>>

abstract: |
  Understanding how humans interact with articifical intelligence through experiments becomes increasingly relevant. However, human-AI interaction researchers lack the appropriate tools to conduct power and data analyses for the required complex study designs. In this work, we provide a tutorial on how to run customized power analyses using data simulation based on Generalized Linear Mixed Models (GLMMs). By providing code in a case study, we equip human-AI interaction researchers with the tools to simulate their own data and run analyses based on GLMMs. We discuss implications for human-AI interaction research and give an outlook.

@@

<!-- General intro, relevance of Human AI Interaction experiments, need for well-powered and analyzed experimental research in this field -->
As the integration of artificial intelligence (AI) into our daily lives continues to grow, it has become increasingly important to study how humans interact with these intelligent systems. 
This is particularly important as AI becomes more sophisticated and is used in decision-making processes, for instance in medicine, that have significant impacts on individuals and society as a whole. 
Therefore, thorough experiments are needed to study human-AI interaction to understand how people perceive and respond to these novel technologies. 
When conducting experimental research in this emerging field, it is essential to determine an appropriate sample size to ensure that the results obtained are both statistically significant and reliable, and use appropriate statistical models to analyze the data.

<!-- Some basics on statistical power, power analysis and why it is important -->
In empirical research relying on hypothesis testing, the most common strategy to determine appropriate samples size is based on statistical power. 
Statistical power is defined as the probability that a hypothesis test of interest turns out significant, when analyzing repeated samples from a population with a true effect of some pre-specified size.
Less formally, power is also described as the probability that a hypothesis test accepts the alternative hypothesis $H_1$ if $H_1$ is indeed true.
In line with this, power analysis represents the act of calculating the statistical power for a given true effect and sample size.

<!-- Power simulations are hard! (especially for GLMMs) -->
Hereby, for a desired statistical power and an assumed true effect, the required sample size to find such an effect can be calculated.
Thereby, power analysis offers a valuable solution by allowing researchers to estimate the appropriate sample sizes needed to achieve sufficient statistical power for reliable results.
While user-friendly software is available for simple statistical models like t-tests, ANOVA, and linear regression, the increasing use of (generalized) linear mixed models (GLMMs; also called multilevel models) in human-AI interaction research introduces greater complexity to power analysis.
For GLMMs, analytic options to derive sample sizes are not available for general research designs.
An increasing number of tutorials describe how to perform power simulation for multilevel models.
However, most tutorials discuss only linear mixed models (LMMs) and focus on the most common designs, limiting the guidance available for researchers facing more complex study designs [REF].
Moreover, existing tutorials often rely on heuristics for specifying variance components (e.g. standard deviation of random intercepts) or assume that meta-analytic results or data from pilot studies is available to determine plausible values for all model parameters.
In contrast, simulations are often required, necessitating assumptions for all parameters in the model.
Therefore, running sample size estimations is crucial for designing experiments that can provide meaningful insights into human-AI interaction.
Moreover, many journals and funding agencies now mandate the inclusion of power analysis as part of study protocols and grant proposals, recognizing their significance in ensuring robust and meaningful findings.
By incorporating power analysis into research planning, researchers can enhance the replicability and credibility of their work, ultimately contributing to the advancement of science.

<!-- Research Gap & Summary -->
When planning a human-AI interaction experiment with complexities such as the use of GLMMs, non-standard designs, and missing by design conditions, the process of power simulations becomes challenging.
Currently available software packages for power simulations do not include common designs in human-AI interaction research, making it necessary to build data simulations tailored specifically to the study design.
Unfortunately, there are no general recommendations that can be applied universally in such cases.
Instead, researchers must acquire the necessary skills based on the specific application at hand.
In this paper, we present a case study that serves as a practical demonstration of how to perform a data simulation for a concrete study for which no meta-analytic results or conclusive data from pilot studies is available.
Through this case study, we aim to provide guidance and insights into the complexities involved in addressing complex design considerations and obtaining accurate power estimates.

@@

3.  Teaching Power Simulation Skills to Human-AI Interaction Researchers: The ability to conduct bespoke power simulations is a valuable skill that should be taught to Human-AI interaction researchers. By incorporating this training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power and improve the quality of their experimental designs. Equipping Human-AI interaction researchers with the knowledge and tools to perform bespoke power simulations empowers them to make informed decisions and enhance the rigor of their studies.

@@

5.  Recognizing the Value of Power Simulations: Bespoke power simulations are not mere technicalities; they are valuable research contributions that deserve recognition in Human-AI interaction research. They offer insights into the reliability and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power. The importance of bespoke power simulations can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports.

@@

In our case study, we have performed power simulations based on a single set of parameter values that reflect our assumptions of an expected effect size.
Instead of extracting this expected effect size from meta-analyses or pilot data which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge which can be considered a vague theoretical model about the data-generating process.
When sample sizes are chosen based on the results of our power simulations, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present.
However, if the true effect is indeed smaller, power will be lower the study might not be sufficiently informative.
A common, more conservative strategy for sample size justification is to perform power analysis for a smallest effect size of interest (SESOI). 
An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero.
For strategies on the even more difficult task of specifying a plausible SESOI as well as a thorough discussion of various topics concerning power analysis, see [@lakensImprovingYourStatistical2022].
When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power for different effect sizes in what is called *sensitivity power analysis*.
By simulating power for different effect sizes (in addition to the different number of subjects and items), one can make sure that power would still be sufficient to detect smaller effect sizes than our expected effect, or at least get an impression how strongly power depends on the size of the true effect.
In simple study designs, it is possible to perform sensitivity power analysis based on a single standardized effect size (e.g. analyse power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8).
However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions.
In this setting, sensitivity power analysis would require to manually specify additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question.
Power could then be simulated for several of these scenarios (across different number of subjects and items as considered earlier).

@@

Power simulations play a critical role in human-AI interaction research, allowing researchers to tailor power estimation to the unique aspects of their experiments. The skills required to perform these simulations should be taught to Human-AI interaction researchers, fostering a deeper understanding of statistical power and enhancing research design. It is essential to bridge the gap between perceived and actual effort associated with power simulations and recognize their value as research contributions. By integrating bespoke power simulations with open science and preregistration practices, researchers can improve the robustness and transparency of their findings, advancing the field.
